Two applications of logic programming to Coq

### 2020-11-05 papers/types2020/sec3.tex
proof system fixed

% \todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
\begin{footnotesize}
\begin{figure}[t]
  %
  \renewcommand{\Gamma}{}
    \newcommand{\bc}[5]{#1#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi\Gamma\vdash A : s}
{%(A,s) \in \mathcal{A}
  \mbox{call to Coq?}
}
% \qquad
% \infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
% {\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
\qquad
\infer{\XXi_1\Gamma\vdash k\ l : A}
      {\bc{\XXi_2}{\Gamma} l D A \quad k : D \in \Pscr_A\qquad \unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ ::\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
\end{figure}
\begin{figure}
\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

Figure~\ref{fig:augmented} contains an implicitly focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
%
It is a sequent calculus inspired by the calculus for
proof search in Pure Type Systems introduced in \cite{LengrandDM06}, based in
 turn on ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a term
language that includes terms and lists of terms, and two distinct typing
judgments for the two categories. In our case, where proof search steps are
viewed as computation, the term judgment corresponds to a goal-reduction
step while the list judgment corresponds to a backchaining step. This coincides
with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.

@@

Since we are only dealing with types that correspond to Horn clauses, and
atomic types are inductively defined, we do not need a $\forall$ rule on the right, although the proof theory would gladly allow it. This means, at the term level, that we
do not have variables in our grammar of terms. Terms are always
applied to a (possibly empty) list of arguments% , and as is usual in Pure Type
% Systems we have the same grammar for terms and types
.
%, as shown in figure~\ref{fig:terms}.
% \todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
% %\begin{figure}
% \begin{align*}
% t,u ::= & k\ l \;|\; \forall x : t.\, u \\
% l ::= & [] \;|\; (t\ ::\ l)
% \end{align*}
% % \caption{Grammar of terms}
% % \label{fig:terms}
% % \end{figure}

\todo{Lookup Coq rules for unfold rule}
The proof system is parameterized by  external set of inductive definitions available in
the (implicit) global context: we denote by $\Pscr_A$ the set of type declarations for the
constructors of type $A$, of the form $k : D$ where $D$ is of the form
$\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We denote
by $\mathcal{A}$ the set of arity declaration for the inductive types.

@@

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog

>>>>

% \todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
\begin{footnotesize}
\begin{figure}[t]
  %
%  \renewcommand{\Gamma}{}
    \newcommand{\bc}[5]{#1#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi\vdash A : s}
{%(A,s) \in \mathcal{A}
  {E[] \vdash_{CIC} A  : s}
}
% \qquad
% \infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
% {\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
\qquad
\infer{\XXi_1\vdash k\ l : A}
      {\mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad (\mathit{head} \ A : T)\in \Gamma_I\quad  k : D \in \Gamma_C\quad \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{}{(t\ ::\ l)} {\forall x:B.\, D} A}
  {\XXi_1\vdash t : A\qquad \bc{\XXi_2}{}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
\end{figure}
\begin{figure}
\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

Figure~\ref{fig:augmented} contains an implicitly focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
%
It is a sequent calculus inspired by the calculus for
proof search in Pure Type Systems introduced in \cite{LengrandDM06}, based in
 turn on ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a term
language that includes terms and lists of terms, and two distinct typing
judgments for the two categories. In our case, where proof search steps are
viewed as computation, the term judgment $\XXi\vdash t \colon A$ corresponds to a goal-reduction
step while the judgment $\XXi\vdash l \colon D \Downarrow A$ corresponds to a backchaining step. This coincides
with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.

The proof system is parameterized by Coq' global environment $E$, which
here can be thought to consist of a set of inductive definitions,
which following Coq's reference manual we denote as
$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  The first rule delegates
to Coq's type checking enforcing the well-sortedness of such types. We
do not have a \emph{local context} since we are only dealing with types
that correspond to Horn clauses, and atomic types are inductively
defined. In fact, we do not have a $\forall$ rule on the right,
although the proof theory would gladly allow it. This means, at the
term level, that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.

The \emph{unfold} rule is the equivalent in our simplified setting of
a \emph{decide} rule in a explicitly focused system. From a logic
programming viewpoint given an atom, it selects a clause on which to
back-chain on: we look-up the costructors from an \lsti{Inductive}
definition from the global enviroment, one that matches the atom we
atom we wish to bachchain on and then call the latter judgment. The rules for backchaining are standard.

@@

  % , and as is usual in Pure Type
% Systems we have the same grammar for terms and types
%
%, as shown in figure~\ref{fig:terms}.
% \todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
% %\begin{figure}
% \begin{align*}
% t,u ::= & k\ l \;|\; \forall x : t.\, u \\
% l ::= & [] \;|\; (t\ ::\ l)
% \end{align*}
% % \caption{Grammar of terms}
% % \label{fig:terms}
% % \end{figure}

% we denote by $\Pscr_A$ the
% set of type declarations for the constructors of type $A$, of the form
% $k : D$ where $D$ is of the form
% $\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We
% denote by $\mathcal{A}$ the set of arity declaration for the inductive
% types.

@@

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog CIC
